## Integration to Seurat 

This vignette is targeted at Seurat users that want to use fcoex. 

A more in depth explanation of each fcoex step is available in the general vignette, which is based on bioconductor packages.

Another popular framework for single-cell analysis is Seurat.
Let's see how to integrate it with fcoex. 


```{r Running Seurat pipeline, warning=FALSE}
library(Seurat)
library(fcoex)
library(ggplot2)

data(pbmc_small)

exprs <- data.frame(GetAssayData(pbmc_small))
target <- Idents(pbmc_small)

fc <- new_fcoex(data.frame(exprs),target)
fc <- discretize(fc)
fc <- find_cbf_modules(fc,n_genes = 70, verbose = FALSE, is_parallel = FALSE)
fc <- get_nets(fc)
```

Once again we can use the Reactome pathways from the package CEMiTool to exemplify how to run an enrichment for human genes.
It is likely that some modules will not have any enrichment, leading to messages of the type "no gene can be mapped.". 
That is not a problem.
If you are not working with human genes, you can just skip this part.

```{r}
gmt_fname <- system.file("extdata", "pathways.gmt", package = "CEMiTool")
gmt_in <- pathwayPCA::read_gmt(gmt_fname)
fc <- mod_ora(fc, gmt_in)

# In Seurat's sample data, pbmc small, no enrichments are found. 
# That is way plot_ora is commented out.

#fc <- plot_ora(fc)
```

```{r Saving Seurat plots, eval = FALSE}
save_plots(name = "fcoex_vignette_Seurat", fc, force = TRUE, directory = "./Plots")
```


### Plotting clusters with  Seurat
Now let's recluster fcoex and visualize the new clusters via the UMAP saved in the Seurat object.
```{r Plotting and saving reclusters,  eval = FALSE}

fc <- recluster(fc) 

file_name <- "pbmc3k_recluster_plots.pdf"
directory <- "./Plots/"

pbmc_small <- RunUMAP(pbmc_small, dims = 1:10)

pdf(paste0(directory,file_name), width = 3, height = 3)
print(DimPlot(pbmc_small))
for (i in names(module_genes(fc))){
  Idents(pbmc_small ) <-   idents(fc)[[i]]
  mod_name <- paste0("M", which(names(idents(fc)) == i), " (", i,")")

  plot2 <- DimPlot(pbmc_small, reduction = 'umap', cols = c("darkgreen", "dodgerblue3")) +
    ggtitle(mod_name) 
    print(plot2)
}
dev.off()
```


The clusters generate by fcoex match possible matches different Seurat clusters. Looking at the HN clusters:
M1 matches cluster 1 (likely monocytes), 
M2 and M4 match  clusters 1 and 2 (likely APCs, B + monocytes),
M5 matches cluster 2 (likeky B)
M7 maches a subset of cluster 0, and as it includes granzymes and granulolysins, this subset of 0 is likely cytotoxic cells (either NK or CD8)

Let's just take a look at the M2 individually:
  
  
  ```{r}
fc <- recluster(fc) 
pbmc_small <- RunUMAP(pbmc_small, dims = 1:10)



Idents(pbmc_small ) <-   target
p1 <- DimPlot(pbmc_small)
Idents(pbmc_small ) <-   idents(fc)[["HLA-DRB1"]]

mod_name <- paste0("M", which(names(idents(fc)) == "HLA-DRB1"), " (", "HLA-DRB1",")")

p2 <- DimPlot(pbmc_small, cols = c("darkgreen", "dodgerblue3")) +
  ggtitle(mod_name) 

# CD79A is a marker of B cells
CD79A <- FeaturePlot(pbmc_small, "CD79A")

# AIF1 is a marker of monocytes
AIF1 <- FeaturePlot(pbmc_small, "AIF1")


library(gridExtra)
grid.arrange(p1, p2, p3,p4, ncol = 2)

```


### Detecting anticorrelated genes in the modules

As the dataset used here is a small subset of the original, some cells might be in unexpected clusters.

The modules capture also negative correlations. These can be especially interesting, as they point to complementary cell types.
Let's look if the module genes are enriched in the cluster HP ( header positive, in blue) or HN (header negative, in green).


```{r, message=FALSE}

for (i in names(module_genes(fc))){
Idents(pbmc_small ) <-   fc@mod_idents[[i]]

# This bit prints which gene in the module belongs to each cluster. 
# HP is the header-positive cluster (containing SOX19A), HN is the header negative cluster (not containing SOX19A)
# The "features = fc@module_list[[i]]" parameter tells Seurat to compare only the genes in the module "i"
# By removing this parameter, you can potentially expand the list that was retrieved originally by fcoex

# Run only for module genes:
module_genes_in_clusters <- FindAllMarkers(pbmc_small, logfc.threshold = 1, only.pos = TRUE, features = fc@module_list[[i]] )

if("HN" %in% module_genes_in_clusters$cluster){
module_genes_in_clusters$module = i
message(paste0("anticorrelated genes found for module ", i))
print(module_genes_in_clusters) 
}
}
```

There seems that only in the module HLA-DRB1 we have negative correlations. Let's visualize then

```{r}
TUBB1 <- FeaturePlot(pbmc_small, "TUBB1")
DRB1 <-  FeaturePlot(pbmc_small, "HLA-DRB1")


library(gridExtra)
grid.arrange(p1, p2, TUBB1, DRB1, ncol = 2)

```


